---
title: "Riding the Yield Curve (Group 9)"
authors: "F. Zhao, L.C. Lacombe, R.J. Mungroop, W.F. Staring"
case-designer: "Michiel Hopman"
date: "2025-01-31"
---


================================
PART 1: DATABASE AND CODE SET UP
================================
```{r, warning=F}
# Load all necessary libraries
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(corrplot)
library(randomForest)

# Clear Environment
rm(list = ls())
graphics.off()

# Set seed for reproduction
set.seed(1) 

# Set name of dataset
file_name <- "Swapcurve 2000-2024 VU.xlsx"

# Set path of dataset
file_path <- "ADD YOU PATH TO THE FILE"

# Read and Load file
data <- read_xlsx(paste0(file_path, file_name))
data$Datum <- as.Date(data$Datum)
```


================================================
PART 2: DATA SCIENCE (CALCULATING PRESENT VALUE)
================================================
```{r PV Computation, warning=F}

# Function that returns date index "years_ahead" years ahead starting from a starting date "date_index" 
jump_years_ahead <- function(date_index, years_ahead) {
  current_date <- data$Datum[date_index]
  next_date <- current_date %m+% years(years_ahead)
  new_index <- which(data$Datum == next_date)
  
  # Go back 1 day until valid date is found
  while (length(new_index) == 0) {
    next_date <- next_date - days(1)
    new_index <- which(data$Datum == next_date)
  }
  
  return(new_index)
}


# Create df to store results
results <- data.frame(
  date = as.Date(character()),
  `1Y` = numeric(),
  `2Y` = numeric(),
  `3Y` = numeric(),
  `4Y` = numeric(),
  `5Y` = numeric())

final_valid_date <- max(data$Datum) - years(5)
final_date_index <- which(data$Datum == final_valid_date)

# Loop through all valid days
for (start_date in 1:final_date_index) { 
  strategy_results <- numeric(5)  # Stores final value of investment for each maturity strategy
  
  for (maturity_strategy in 1:5) {
    current_year <- 1   # Start at T = 1 as R starts indexing at 1
    final_year <- 6
    
    present_value_count <- maturity_strategy - 1
    investment_at_year <- numeric(final_year + present_value_count)   # Stores investment value at given year
    investment_at_year[1] <- 100   # Initial investment
    PV_rates <- numeric(5)    # Stores final year bond rates for PV calculations
    final_investment_value <- 0
  
    while (current_year < final_year) {
      current_date_row <- jump_years_ahead(start_date, current_year - 1)
      current_bond_rate <- data[[paste0(maturity_strategy, "Y")]][current_date_row]
      
      final_maturity_year <- current_year + maturity_strategy
      
      # If bond will mature past investment horizon get PV bond rate at final year 
      if (final_maturity_year > final_year) {
        remaining_years <- final_maturity_year - final_year
        final_date <- jump_years_ahead(start_date, 5)
        
        PV_rates[current_year] <- data[[paste0(remaining_years, "Y")]][final_date]
      }
      
      ## Coupons ##
   
      # Get back full investment after full maturity
      investment_at_year[final_maturity_year] <- investment_at_year[current_year]
      
      # Distribute coupons across the upcoming years
      coupon <- investment_at_year[current_year] * current_bond_rate
      for (years_ahead in 1:maturity_strategy) {
        investment_at_year[current_year + years_ahead] <- investment_at_year[current_year + years_ahead] + coupon
      }
      
      current_year <- current_year + 1
    }
    
    # Add final value from coupons to final investment value
    final_investment_value <- final_investment_value + investment_at_year[final_year]

    ## Present Values (PV) ##

    if (present_value_count != 0) {
      for (PV_i in 1:present_value_count) {
        FV <- investment_at_year[final_year + PV_i]
        r <- PV_rates[final_year - maturity_strategy + PV_i]
        n <- PV_i
        present_value <- FV / (1 + r)^(n)

        final_investment_value <- final_investment_value + present_value
      }
    }

    # Store final investment value of maturity
    strategy_results[maturity_strategy] <- final_investment_value
  }
  
  # Store all investment strategy results
  results[nrow(results) + 1, ] <- c(
    data$Datum[start_date],
    strategy_results[1],
    strategy_results[2],
    strategy_results[3],
    strategy_results[4],
    strategy_results[5])
}
```


==============================
PART 3: DESCRIPTIVE STATISTICS
==============================
```{r, dpi=300, fig.width=10, fig.height=6}
## Plot interest rates over time ##

plot(data$Datum, data$`1Y`,
     xlab = "Date",
     ylab = "Interest Rate",
     main = "Daily Fixed Rates",
     type = "l",
     lwd = 1,
     col = "orange")
lines(data$Datum, data$`2Y`, lwd = 1, col = "red")
lines(data$Datum, data$`3Y`, lwd = 1, col = "green")
lines(data$Datum, data$`4Y`, lwd = 1, col = "skyblue")
lines(data$Datum, data$`5Y`, lwd = 1, col = "purple")
legend("bottomleft",
       legend = c("1Y", "2Y", "3Y", "4Y", "5Y"),
       fill = c("orange", "red", "green", "skyblue", "purple"))


# Modify column names for plots (X1Y -> 1Y)
colnames(results)[2:6] <- c("1Y", "2Y", "3Y", "4Y", "5Y")

## Plot strategy results over time ##

# Plot the 1Y strategy
plot(results$date, results$`1Y`,
     xlab = "Date",
     ylab = "Strategy Results",
     main = "Daily Strategy Results",
     type = "l",
     col = "orange",   # Color for the first line
     lwd = 1,
     ylim = c(90, 150)  # Set the y-axis range from 90 to 150
)

# Add other strategies
lines(results$date, results$`2Y`, col = "red", lwd = 1)
lines(results$date, results$`3Y`, col = "green", lwd = 1)
lines(results$date, results$`4Y`, col = "skyblue", lwd = 1)
lines(results$date, results$`5Y`, col = "purple", lwd = 1)

# Add the legend
legend("topright",
       legend = c("1Y", "2Y", "3Y", "4Y", "5Y"),
       fill = c("orange", "red", "green", "skyblue", "purple"))


### Histograms ###

achmea_red <- "#CC0000"
achmea_red_dark <- "#960000"

# Extract best strategy
max_values <- apply(results[, -1], 1, max)  # Maximum value for each row
max_vars <- apply(results[, -1], 1, function(row) names(results[, -1])[which.max(row)])  # Variable name for max value

# Add to the dataset
results$MaxValue <- max_values
results$MaxVariable <- max_vars

## Histogram of best strategy counts ##

hist_data <- table(results$MaxVariable) 

# Plot the bar plot 
barplot(hist_data,
        main ="Histogram of Best Strategy Count",
        xlab="Strategy",
        ylab="Frequency",
        col = adjustcolor(achmea_red, alpha.f = 0.7),
        border = achmea_red_dark)

## Histogram of each strategies results ##

# Columns to plot for
strategy_histograms <- c("1Y", "2Y", "3Y", "4Y", "5Y", "MaxValue")

# Loop over each column and create a histogram
for (hist in strategy_histograms) {

  # Create histogram using ggplot
  strategy_histogram <- ggplot(results, aes_string(x = paste0("`", hist, "`"))) +
    geom_histogram(binwidth = 2, fill = achmea_red, color = achmea_red_dark, alpha = 0.7) +
    labs(title = paste("Histogram of", hist),x = "Investment Value", y = "Frequency") +
    xlim(85, 135) +  # Set x-axis range
    ylim(0, 750) +   # Set y-axis range
    theme_minimal()

  # Display the plot
  print(strategy_histogram)
}

## Box plots for each strategy and best strategy ##

boxplot_data <- results[, c("1Y", "2Y", "3Y", "4Y", "5Y", "MaxValue")]

boxplot(boxplot_data,
        names = c("1Y", "2Y", "3Y", "4Y", "5Y", "MaxValue"),
        col = adjustcolor(achmea_red, alpha.f = 0.7),
        border = achmea_red_dark,
        main = "Boxplot of Investment Strategies",
        xlab = "Strategy",
        ylab = "Investment Value",
        lwd = 1.5)

# Add grid lines
grid(nx = 0, ny = NULL, lty = "dotted", col = adjustcolor("gray", alpha.f = 0.5), lwd = 1.5)


## Calculate and print summary statistics ##

cat("\n")
summary(results[, 2:7])

```


================
PART 4: MODELING
================

Random Forest (RF) Out of Sample Back Testing (COMPUTATION)
```{r out-of-sample backtesting}

# Create df for modeling data
full_train_data <- data.frame(
  Datum = data$Datum,
  Slope = data$`5Y` - data$`1Y`,
  Level = rowMeans(data[, c("1Y", "2Y", "3Y", "4Y", "5Y")])
)


full_train_data <- cbind(
  full_train_data[1:nrow(results),], # Ensure columns have the same length
  results[2:ncol(results)] # PV results
  )

# Convert MaxVariable to type factor (for model)
full_train_data$MaxVariable <- as.factor(full_train_data$MaxVariable)


## BACK TESTING ##

# Around 12 years
window_size <- 3024   

# Number of predictions per iteration (window of 10 days)
n_predictions <- 10

# Gap between training data and prediction data (essential to model real life)
five_year_gap <- 1260   # Around 5 years

# Df to store the model predictions
rolling_results <- data.frame(
  historical_best = character(),
  predicted_best = character(),
  stringsAsFactors = FALSE
)

# FOR loop to roll through all windows

total_windows <- nrow(full_train_data) - window_size - five_year_gap - n_predictions

for (i in 1:total_windows) {
  
  # Window of data used to train the model
  train_data_window <- droplevels(full_train_data[i:(i + window_size - 1), ])
  
  # Skip iteration if there is only one used strategy in train_data
  # This step is necessary for the ML models to work
  if(length(levels(train_data_window$MaxVariable)) < 2) next
        
  # Window of days on which we make predictions
  prediction_window <- full_train_data[
    (i + window_size + five_year_gap):
    (i + window_size + five_year_gap + n_predictions - 1), ]
  
  # Train Random Forest model
  rf_model <- randomForest(MaxVariable ~    # Target Variable
                           Slope + Level,   # Predictive Variables
                           data = train_data_window,
                           ntree = 64)
  
  # Extract predictive power of each variable
  importance_vals <- importance(rf_model, type = 2)
  gini_level <- importance_vals["Level", "MeanDecreaseGini"]
  gini_slope <- importance_vals["Slope", "MeanDecreaseGini"]
  
  
  # Use trained model to make predictions
  rf_prediction <- predict(rf_model, 
                        newdata = prediction_window)
  
  
  # Append this iterations RF results
  rolling_results <- rbind(rolling_results,
      data.frame(
      historical_best = as.character(prediction_window$MaxVariable),
      predicted_best = rf_prediction,
      gini_level = gini_level,
      gini_slope = gini_slope,
      stringsAsFactors = FALSE
    )
  )
}

```


RF Out of Sample Back Testing (RESULTS)
```{r backtesting results}

# Accuracy for Random Forest back testing models
accuracies <- matrix(
  data = c(
  mean(rolling_results$historical_best == rolling_results$predicted_best)),
  byrow = T,
  nrow = 1,
  ncol = 1
)
rownames(accuracies) <- c("W_E=3024")
colnames(accuracies) <- c("ntree = 64")

# Display accuracies
cat("Accuracies for each model:\n")
print(accuracies)

# Function for confusion matrices
confusion_matrix <- function(prediction, reference){
  confusion <- matrix(data = 0, nrow = 5, ncol = 5)
  rownames(confusion) <- colnames(confusion) <- c("1Y", "2Y", "3Y", "4Y", "5Y")
  
  for(i in 1:length(prediction)){
    predicted <- as.integer(substr(prediction[i], 1, 1))
    actual <- as.integer(substr(reference[i], 1, 1))
    
    confusion[actual, predicted] <- confusion[actual, predicted] + 1
  }
  
  return(confusion)
}

# Display confusion matrix
cat("\nConfusion matrix of Random Forest backtest:\n")
confusion_matrix(
  factor(rolling_results$predicted_best), 
  factor(rolling_results$historical_best)
  )

# Average predictive power of each variable
avg_gini_level <- mean(rolling_results$gini_level, na.rm = TRUE)
avg_gini_slope <- mean(rolling_results$gini_slope, na.rm = TRUE)

cat("\nAverage Gini Importance:\n")
cat("Level:", avg_gini_level, "\n")
cat("Slope:", avg_gini_slope, "\n")
```


RF Out of Sample Back Testing (COMPUTATION & RESULTS)
```{r in-sample testing}

# Fit model on all available data
rf_in_sample <- randomForest(MaxVariable ~ 
                                Slope + Level,
                                data = full_train_data,
                                ntree = 64)

## Predict all available data
rf_predictions_in_sample <- predict(rf_in_sample, newdata = full_train_data)

# Confusion matrix
confusion_matrix(
  factor(full_train_data$MaxVariable),
  factor(rf_predictions_in_sample)
)

# Accuracy
accuracy <- mean(full_train_data$MaxVariable == rf_predictions_in_sample)

cat("\nIn-sample accuracy:", accuracy)


# Predictive power of each variable

cat("\nAverage Gini Importance:\n")
print(importance(rf_in_sample))


```


Evaluation of ML Models Predictive Variables (Slope vs Level)
```{r, dpi=300, fig.width=10, fig.height=6}
# Scatterplot of Slope vs Level
ggplot(full_train_data, aes(x = Slope, y = Level, color = MaxVariable)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("orange", "red", "green", "skyblue", "purple")) +
  theme_minimal() + 
  labs(
    title = "Scatterplot of Best Strategy by Slope vs Level",
    x = "Slope",
    y = "Level",
    color = "Best Strategy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13)
  )

```


Evaluation of ML Models Predictive Variables (Slope/Level vs Present Value)
```{r, dpi=300, fig.width=10, fig.height=6}

# Example: Reshape the data into a long format for easy plotting
results_long <- full_train_data %>%
  select(Slope, Level, '1Y', '2Y', '3Y', '4Y', '5Y') %>%
  pivot_longer(cols = '1Y':'5Y', names_to = "Strategy", values_to = "PresentValue")

# Scatter plot: Correlation between Slope and Present Value
ggplot(results_long, aes(x = Slope, y = PresentValue, color = Strategy)) +
  geom_point(alpha = 0.5) + 
  labs(
    title = "Correlation Between Slope and Present Value for Each Strategy",
    x = "Slope",
    y = "Present Value",
    color = "Strategy"
  ) +
  scale_color_manual(values = c("orange", "red", "green", "skyblue", "purple")) +
  theme_minimal()

# Scatter plot: Correlation between Level and Present Value
ggplot(results_long, aes(x = Level, y = PresentValue, color = Strategy)) +
  geom_point(alpha = 0.5) + 
  labs(
    title = "Correlation Between Level and Present Value for Each Strategy",
    x = "Level",
    y = "Present Value",
    color = "Strategy"
  ) +
  scale_color_manual(values = c("orange", "red", "green", "skyblue", "purple")) +
  theme_minimal()

strategies <- c('1Y', '2Y', '3Y', '4Y', '5Y')

# Calculate correlations
slope_cor <- sapply(strategies, function(s) cor(full_train_data$Slope, full_train_data[[s]]))
level_cor <- sapply(strategies, function(s) cor(full_train_data$Level, full_train_data[[s]]))

# Print correlations
cat("Slope Correlation:\n")
for (i in seq_along(strategies)) {
  cat(sprintf("%s: %.4f\n", strategies[i], slope_cor[i]))
}

cat("\nLevel Correlation:\n")
for (i in seq_along(strategies)) {
  cat(sprintf("%s: %.4f\n", strategies[i], level_cor[i]))
}

```


RF vs Historic Statistics (HS) Models Performance Evaluation (COMPUTATION)
```{r Strategy Results Functions }

# Function that uses Random Forest Model with rolling window
RF_prediction <- function(predict_date_index, i) {
  
  # Rolling Window
  train_window <- full_train_data[i:(i + window_size), ]

  # Train model
  model <- randomForest(MaxVariable ~ Slope + Level,
                       data = train_window,
                       ntree = 64)

  # Make prediction on single (current day)
  prediction <- as.character(predict(model,
                                      newdata = full_train_data[predict_date_index,]))

  # Get prediction value
  prediction_value <- full_train_data[predict_date_index, prediction]
  
  best_strat <- as.character(full_train_data[predict_date_index, "MaxVariable"])
  best_value <- full_train_data[predict_date_index, best_strat]

  return(data.frame(date_index = predict_date_index,
                    prediction = prediction,
                    prediction_value = prediction_value,
                    best_strat = best_strat,
                    best_value = best_value))
}


# Function that makes predictions using the historic best strategy (increasing window)
historic_prediction <- function(predict_date_index, i) {

  # Increasing window size that starts at same size as RF
  historical_window <- full_train_data[1:(i + window_size), ]

  # Find the most frequent value best value and set as prediction
  counts <- table(historical_window$MaxVariable)
  prediction <- names(counts)[which.max(counts)]

  # Get prediction value
  prediction_value <- full_train_data[predict_date_index, prediction]

  best_strat <- as.character(full_train_data[predict_date_index, "MaxVariable"])
  best_value <- full_train_data[predict_date_index, best_strat]
  

  return(data.frame(date_index = predict_date_index,
                    prediction = prediction,
                    prediction_value = prediction_value,
                    best_strat = best_strat,
                    best_value = best_value))
}

```


RF vs HS Models Performance Evaluation (Function Call)
```{r Calculate Strategy Results }
# Create data frames to store results
RF_results <- data.frame(
  date_index = integer(),
  prediction = character(),
  prediction_value = numeric(),
  best_strat = character(),
  best_value = numeric())

historic_results <- data.frame(
  date_index = integer(),
  prediction = character(),
  prediction_value = numeric(),
  best_strat = character(),
  best_value = numeric())

# Remove 5 years from data to recreate real life conditions
investement_strat_data <- full_train_data[1:(nrow(full_train_data) - five_year_gap), ]


## Loop through all possible days that can be predicted and find model predictions
max_predictions <- nrow(full_train_data) - five_year_gap - window_size

for (i in 1:max_predictions) {
  pred_date_index <- i + window_size + five_year_gap

  RF_best_strat <- RF_prediction(pred_date_index, i)
  RF_results <- rbind(RF_results, RF_best_strat)

  historic_best_strat <- historic_prediction(pred_date_index, i) 
  historic_results <- rbind(historic_results, historic_best_strat)
}

```


RF vs HS Models Performance Evaluation (RESULTS)
```{r Strategy Comparison }
# Add profit/loss columns
RF_results$predicted_profit <- RF_results$prediction_value - 100
RF_results$best_profit <- RF_results$best_value - 100
RF_results$lost_profit <- ifelse(
  RF_results$prediction != RF_results$best_strat, 
  RF_results$best_profit - RF_results$predicted_profit, 0)

historic_results$predicted_profit <- historic_results$prediction_value - 100
historic_results$best_profit <- historic_results$best_value - 100
historic_results$lost_profit <- ifelse(
  historic_results$prediction != historic_results$best_strat, 
  historic_results$best_profit - historic_results$predicted_profit, 0)

# Accuracy: Percentage of days where predicted strategy matched best strategy
RF_accuracy <- mean(RF_results$prediction == RF_results$best_strat)
historic_accuracy <- mean(historic_results$prediction == historic_results$best_strat)

# Profit Metrics (dollar terms)
RF_avg_profit <- mean(RF_results$predicted_profit)
historic_avg_profit <- mean(historic_results$predicted_profit)

RF_net_returns <- sum(RF_results$predicted_profit)
historic_net_returns <- sum(historic_results$predicted_profit)

# Opportunity Cost Metrics
RF_total_lost <- sum(RF_results$lost_profit)
historic_total_lost <- sum(historic_results$lost_profit)

RF_avg_lost <- mean(RF_results$lost_profit)
historic_avg_lost <- mean(historic_results$lost_profit)

# Risk Metrics (Percentage Returns)
RF_returns <- RF_results$predicted_profit / 100  
historic_returns <- historic_results$predicted_profit / 100

# Sharpe Ratio
risk_free_rate <- 0
RF_sharpe <- (mean(RF_returns) - risk_free_rate) / sd(RF_returns)
historic_sharpe <- (mean(historic_returns) - risk_free_rate) / sd(historic_returns)

# Standard deviation of returns
RF_volatility <- sd(RF_returns)
historic_volatility <- sd(historic_returns)

# Function to compute Value-at-Risk and Expected Shortfall calculations
compute_var_ES <- function(returns, confidence) {
  var_value <- quantile(returns, probs = 1 - confidence, na.rm = TRUE)
  ES_value <- mean(returns[returns <= var_value], na.rm = TRUE)
  
  return(c(var_value, ES_value))
}

# 1%, 5%, 10% confidence levels for VAR and ES
RF_VaR_1 <- compute_var_ES(RF_returns, 0.99) 
RF_VaR_5 <- compute_var_ES(RF_returns, 0.95)
RF_VaR_10 <- compute_var_ES(RF_returns, 0.90)

historic_VaR_1 <- compute_var_ES(historic_returns, 0.99)
historic_VaR_5 <- compute_var_ES(historic_returns, 0.95)
historic_VaR_10 <- compute_var_ES(historic_returns, 0.90)

# Function to format table results
format_value <- function(x, digits = 4) {
  if(is.infinite(x)) return("Inf")
  return(format(round(x, digits), scientific = FALSE, nsmall = digits, trim = TRUE))
}

# Create comparison data frame
results_table <- data.frame(
  Metric = c(
    "Model Accuracy", 
    "Average Profit ($)", 
    "Total Profit ($)", 
    "Total Lost Profit ($)",
    "Average Lost Profit ($)", 
    "Sharpe Ratio", 
    "Volatility", 
    "VaR 1%", 
    "VaR 5%", 
    "VaR 10%", 
    "ES 1%", 
    "ES 5%", 
    "ES 10%"
  ),
  RF = sapply(
    c(
      RF_accuracy, 
      RF_avg_profit, 
      RF_net_returns, 
      RF_total_lost,
      RF_avg_lost, 
      RF_sharpe, 
      RF_volatility, 
      RF_VaR_1[1], RF_VaR_5[1], RF_VaR_10[1], 
      RF_VaR_1[2], RF_VaR_5[2], RF_VaR_10[2]
    ), 
    format_value
  ),
  Historic = sapply(
    c(
      historic_accuracy, 
      historic_avg_profit, 
      historic_net_returns,
      historic_total_lost, 
      historic_avg_lost,
      historic_sharpe,
      historic_volatility, 
      historic_VaR_1[1], historic_VaR_5[1], historic_VaR_10[1], 
      historic_VaR_1[2], historic_VaR_5[2], historic_VaR_10[2]
    ), 
    format_value
  ),
  Best = c(
    # Accuracy
    ifelse(RF_accuracy >= historic_accuracy, "RF", "Historic"),
    # Average Profit ($)
    ifelse(RF_avg_profit >= historic_avg_profit, "RF", "Historic"),
    # Total Profit ($)
    ifelse(RF_net_returns >= historic_net_returns, "RF", "Historic"),
    # Total Lost Profit ($) (lower is better)
    ifelse(RF_total_lost <= historic_total_lost, "RF", "Historic"),
    # Average Lost Profit ($) (lower is better)
    ifelse(RF_avg_lost <= historic_avg_lost, "RF", "Historic"),
    # Sharpe Ratio (higher is better)
    ifelse(RF_sharpe >= historic_sharpe, "RF", "Historic"),
    # Volatility (lower is better)
    ifelse(RF_volatility <= historic_volatility, "RF", "Historic"),
    # VaR/ES comparisons (less negative/more positive is better)
    ifelse(RF_VaR_1[1] >= historic_VaR_1[1], "RF", "Historic"),
    ifelse(RF_VaR_5[1] >= historic_VaR_5[1], "RF", "Historic"),
    ifelse(RF_VaR_10[1] >= historic_VaR_10[1], "RF", "Historic"),
    ifelse(RF_VaR_1[2] >= historic_VaR_1[2], "RF", "Historic"),
    ifelse(RF_VaR_5[2] >= historic_VaR_5[2], "RF", "Historic"),
    ifelse(RF_VaR_10[2] >= historic_VaR_10[2], "RF", "Historic")
  )
)

# Print formatted results
print(results_table, row.names = FALSE)


print(sum(historic_results$best_profit))
```

